### 2024-02-20 paper_sisc_first_revision/localpsf.tex
one readthrough complete

The PSF approximation $\mathbf{A}_H$, and therefore $\mathbf{A}_H + \mathbf{R}_H$, may be non-symmetric and indefinite because of approximation error. 
This is undesirable because symmetry and positive semi-definiteness are important properties which should be preserved if possible. Also, lacking these properties may prevent one from using highly effective algorithms to perform further operations involving $\mathbf{A}_H + \mathbf{R}_H$, such as using $\mathbf{A}_H + \mathbf{R}_H$ as a preconditioner in the conjugate gradient method.

@@

In summary, computing all the necessary ingredients to evaluate kernel entries of the PSF approximation requires a handful of operator applications (e.g., $6+\nbatch$ operator applications in two dimensions, or $10+\nbatch$ operator applications in three dimensions, with $\nbatch$ typically in the range 1--25), plus comparatively cheap additional overhead costs, most notably performing ellipsoid intersection tests while choosing sample point batches. Once these ingredients are computed, no more operator applications (and thus PDE solves) are required, and approximate kernel entries can be evaluated rapidly. Constructing the H-matrix from kernel entries requires a number of elementary operations that scales polylog linearly in $N$. Using the H-matrix to perform linear algebra operations also scales polylog linearly in $N$, though the details of these costs depend heavily on the type of H-matrix and operation being performed.

@@

Next, we build PSF (1), PSF (5), and PSF (25) preconditioners based on the Gauss-Newton Hessian evaluated at the converged solution $\mathbf{\basalfriction}$. We use PCG to solve a linear system with the Hessian as the coefficient operator and a right hand vector with random independent and identically distributed (i.i.d.) entries drawn from the standard Gaussian distribution. In Figure~\ref{fig:krylov_convergence} (left) we compare the convergence of PCG for solving this linear system using the PSF (1), PSF (5), PSF (25), REG, and NONE preconditioners. PCG converges fastest with the PSF-based preconditioners, with PSF (25) converging fastest, followed by PSF (5), followed by PSF (1), as expected. 
In Figure~\ref{fig:krylov_convergence} (right) we show the generalized eigenvalues for the generalized eigenvalue problem
$
\mathbf{H} \mathbf{u} = \lambda \mathbf{\preconditioner} \mathbf{u}.
$ 
The matrix $\mathbf{\preconditioner}$ is one of the PSF (1), PSF (5), or PSF (25) Gauss-Newton Hessian approximations, the regularization Hessian (REG), or the identity matrix (NONE). With the PSF-based preconditioners, the generalized eigenvalues cluster near one, with more batches yielding better clustering.

@@

\begin{figure}
	\begin{subfigure}{0.52\textwidth}
		\centering
		\includegraphics[scale=0.9]{adv_pcg_convergence.pdf}
	\end{subfigure}
	\begin{subfigure}{0.47\textwidth}
		\centering
		\includegraphics[scale=0.9]{adv_geigs_figure.pdf}
	\end{subfigure}
	\caption{(Advective-diffusive transport) Left: Convergence history for solving $\mathbf{H} \mathbf{x} = \mathbf{b}$ using PCG, where $\mathbf{b}$ has i.i.d.\ random entries drawn from the standard Gaussian distribution. Right: Generalized eigenvalues for generalized eigenvalue problem $\mathbf{H} \mathbf{u}_{k}=\lambda_{k} \mathbf{\preconditioner} \mathbf{u}_{k}$. Here, $\mathbf{H}$ is the Hessian and the preconditioner, $\mathbf{\preconditioner}$, is the PSF approximation for 1, 5, or 25 batches (PSF (1), PSF (5), and PSF (25), respectively), the regularization Hessian (REG), or the identity matrix (NONE).}
	\label{fig:adv_krylov_convergence}
\end{figure}

>>>>

The PSF-based approximation $\mathbf{A}_H$, and therefore $\mathbf{A}_H + \mathbf{R}_H$, may be non-symmetric and indefinite because of approximation error. 
This is undesirable because symmetry and positive semi-definiteness are important properties which should be preserved if possible. Also, lacking these properties may prevent one from using highly effective algorithms to perform further operations involving $\mathbf{A}_H + \mathbf{R}_H$, such as using $\mathbf{A}_H + \mathbf{R}_H$ as a preconditioner in the conjugate gradient method.

@@

In summary, computing all the necessary ingredients to evaluate kernel entries of the PSF-based approximation requires a handful of operator applications (e.g., $6+\nbatch$ operator applications in two dimensions, or $10+\nbatch$ operator applications in three dimensions, with $\nbatch$ typically in the range 1--25), plus comparatively cheap additional overhead costs, most notably performing ellipsoid intersection tests while choosing sample point batches. Once these ingredients are computed, no more operator applications (and thus PDE solves) are required, and approximate kernel entries can be evaluated rapidly. Constructing the H-matrix from kernel entries requires a number of elementary operations that scales polylog linearly in $N$. Using the H-matrix to perform linear algebra operations also scales polylog linearly in $N$, though the details of these costs depend heavily on the type of H-matrix and operation being performed.

@@

Next, we build PSF~(1), PSF~(5), and PSF~(25) preconditioners based on the Gauss-Newton Hessian evaluated at the converged solution $\mathbf{\basalfriction}$ (note: $k$ in PSF~($k$) refers to the number of batches; this is not to be confused with the noise levels which range over the same numerical values). We use PCG to solve a linear system with the Hessian as the coefficient operator and a right hand side vector with random independent and identically distributed (i.i.d.) entries drawn from the standard Gaussian distribution. In Figure~\ref{fig:krylov_convergence} (left) we compare the convergence of PCG for solving this linear system using the PSF (1), PSF (5), PSF (25), REG, and NONE preconditioners. PCG converges fastest with the PSF-based preconditioners, with PSF (25) converging fastest, followed by PSF (5), followed by PSF (1), as expected. 
In Figure~\ref{fig:krylov_convergence} (right) we show the generalized eigenvalues for the generalized eigenvalue problem
$
\mathbf{H} \mathbf{u} = \lambda \mathbf{\preconditioner} \mathbf{u}.
$ 
The matrix $\mathbf{\preconditioner}$ is one of the PSF (1), PSF (5), or PSF (25) Gauss-Newton Hessian approximations, the regularization Hessian (REG), or the identity matrix (NONE). With the PSF-based preconditioners, the generalized eigenvalues cluster near one, with more batches yielding better clustering.

@@

\begin{figure}
	\begin{subfigure}{0.52\textwidth}
		\centering
		\includegraphics[scale=0.9]{adv_pcg_convergence.pdf}
	\end{subfigure}
	\begin{subfigure}{0.47\textwidth}
		\centering
		\includegraphics[scale=0.9]{adv_geigs_figure.pdf}
	\end{subfigure}
	\caption{(Advective-diffusive transport) Left: Convergence history for solving $\mathbf{H} \mathbf{x} = \mathbf{b}$ using PCG, where $\mathbf{b}$ has i.i.d.\ random entries drawn from the standard Gaussian distribution. Right: Generalized eigenvalues for generalized eigenvalue problem $\mathbf{H} \mathbf{u}_{k}=\lambda_{k} \mathbf{\preconditioner} \mathbf{u}_{k}$. Here, $\mathbf{H}$ is the Hessian and the preconditioner, $\mathbf{\preconditioner}$, is the PSF-based approximation for 1, 5, or 25 batches (PSF (1), PSF (5), and PSF (25), respectively), the regularization Hessian (REG), or the identity matrix (NONE).}
	\label{fig:adv_krylov_convergence}
\end{figure}
